# Chapter 5: Dimension reduction practice
As usual, let's start with loading the libraries:

```{r, results='hide', message=FALSE}
library(dplyr)
library(GGally)
library(ggplot2)
library(tidyr)
library(caret)
library(MASS)
```

## Dataset description
Load the data we processed in the data wrangling section (actually, I'm using the ready made dataset)
```{r}
data <- read.csv("data/human/human2.txt")
```
Print a summary of the data structure
```{r}
str(data)
```
The dataset includes human development metrics from different countries. The metrics include life expectancy, GNI, education level, and so on.

Plot variable pairs to visualize correlations
```{r}
p <- ggpairs(data, mapping = aes(alpha=0.3), lower = list(combo = wrap("facethist", bins = 20)))
p
```

The strongest correlations can be seen in
- 1
- 2
- 3

Scale the data
```{r}
data.scaled <- data.frame(scale(data))
```

## Analysis

Now we shall run the principal component analysis (PCA) on the data. First with the non-scaled data.

```{r}

```


